\section{What happened to Tali Forth 1?} 

Tali Forth 1\index{Tali Forth 1}, formally just Tali Forth, was my first Forth.
As such, it is fondly remembered as a learning experience. You can still find it
online at GitHub\index{GitHub} at
\href{https://github.com/scotws/TaliForth}{https://github.com/scotws/TaliForth}.
When Tali Forth 2 entered BETA, Tali Forth was discontinued and does not receive
bug fixes either.

\section{Why does Tali Forth take so long to start up?}

After the default kernel string is printed, you'll notice a short pause that
didn't occur with Tali Forth 1. This is because Tali Forth 2 has more words
defined in high-level Forth (see \texttt{forth-words.asm}) than Tali did. The
pause happens because they are being compiled on the fly.

\section{Why `Tali' Forth?}

I like the name, and we're probably not going to have any more kids I can give
it to.

(If it sounds vaguely familiar, you're probably thinking of Tali'Zorah vas
Normandy,\index{vas Normandy, Tali'Zorah} a character in the `Mass
Effect'\index{Mass Effect@\textit{Mass Effect}} universe created by EA/BioWare.  This
software has absolutely nothing to do with either the game or the companies and
neither do I, expect that I've played the games and enjoyed them, though I do
have some issues with \emph{Andromeda}. Like what happened to the quarian ark?)

\section{Who is `Liara'?}\index{Liara Forth|textbf}

Liara Forth is a STC Forth for the big sibling of the 6502, the
65816\index{65816}. Tali Forth 1\index{Tali Forth 1} came first, then I wrote
Liara with that knowledge and learned even more, and now Tali 2 is such much
better for the experience. Oh, and it's another `Mass Effect'\index{Mass
Effect@\textit{Mass Effect}} character.

