% Overview of Tali Forth
% Scot W. Stevenson

% -------------------------------------
\section{Design considerations}

When creating a new Forth, there are a bunch of design decisions to be
made.\footnote{The best introduction to these questions is found in
\href{http://www.bradrodriguez.com/papers/moving1.htm}{\textit{Design Decisions
in the Forth Kernel}} by Brad Rodriguez\index{Rodriguez, Brad}} Spoiler alert:
Tali Forth ended up as a subroutine-threaded variant with a 16-bit cell size and
a dictionary that keeps headers and code separate. If you don't care and just
want to use the program, skip ahead. 

\subsection{Characteristics of the 65c02}

Since this is a bare-metal Forth, the most important consideration is the target
processor. The 65c02 only has one full register, the accumulator
A\index{accumulator}, and two secondary registers X\index{X register} and
Y\index{Y register}. All are 8-bit wide. There are 256 bytes that are more
easily addressable on the Zero Page\index{Zero Page}. A single hardware stack is
used for subroutine jumps. The address bus\index{address bus} is 16 bits wide
for a maximum of 64 KiB of RAM\index{RAM} and \index{ROM} -- for a simple setup,
we should assume 32 KiB of each. 

\subsection{Cell size}

This is relatively easy: The 16 bit address bus\index{address bus} suggests the
cell size should be 16 bits as well. This is still easy enough to realize with a
8-bit MPU, though not as comfortable as working with the 65816\index{65816}, the
65c02's big brother with a 16 bit register size.

\subsection{Threading technique}\index{threading}

A `thread' in Forth is simply a list of addresses of words to be executed. 
There are four basic threading techniques:\footnote{For the 8086 MPU, Guy
Kelly\index{Kelly, Guy} compared various Forth implementations in
1992\cite{kelly92}}

\begin{description}
        \item [Indirect threaded (ITC)] The oldest, original variant, used by FIG
                Forth\index{FIG Forth} for instance. All other versions are
                modifications of this model
        \item [Direct threaded (DTC)] Includes more assembler code to become faster, but
                slightly larger than ITC. 
        \item [Token threaded (TTC)] The reverse of DTC in that it is slower, but uses
                less space than the other Forths. Words are created as a table
                of tokens.
        \item [Subroutine threaded (STC)] This technique converts the words to a simple
                series of \texttt{jsr} combinations. 
\end{description}

Our lack of registers and the goal of creating a simple and easy to understand
Forth makes subroutine threading the most attractive solution. We will try to
mitigate the pain of the 12 cycle cost of every \texttt{jsr/rts} combination by
including a relatively large number of native words\index{native words}. 

\subsection{Register use}

The lack of registers -- and 16 bit registers at that -- becomes apparent when
you realize that Forth classically uses at least four `virtual' registers:

\begin{table}[h !]
        \centering
        \label{tab:registers}
        \begin{tabular}{| c | l |}
                \hline
                W   & Working register\\
                IP  & Interpreter Pointer\\
                DSP & Data Stack Pointer\\
                RSP & Return Stack Pointer\\
                \hline
        \end{tabular}
        \caption{The classic Forth registers}
\end{table}

On a modern processor like the RISC-V\index{RISC-V} RV32I with its 32 registers
of 32 bit each, this wouldn't be a problem. In fact, we'd be trying to figure
out what else we could keep in a register. On the 65c02, at least we get the RSP
for free with the built-in stack pointer. This still leaves three registers. We
cut that number down by one through subroutine threading, which gets rid of the
IP. For the DSP, we use the 65c02's Zero Page\index{Zero Page} indirect
addressing mode with the X register\index{X register}. This leaves W, which we
put on the Zero Page as well. 


\subsection{Data Stack design}

We'll go into greater detail on how the Data Stack works in a later chapter when
we look at the internals. Put briefly, the stack is realized on the Zero
Page\index{Zero Page} for speed. For stability, we provid underflow
checks\index{underflow!detection} in the relevant words, but give the user the
option of stripping it out for native compilation\index{native compilation}.


\subsection{Dictionary structure}

Each Forth word consists of the actual code and the header\index{header} which
keeps the meta-data. Part of this data is the single-linked list of words which
is searched. 

In constrast to Tali Forth 1\index{Tali Forth 1}, which kept the header and body
of the words together, Tali Forth 2 keeps them separate. This lets us play
various tricks with the code to make it more effective.

\subsection{Deeper down the rabbit hole}

This concludes our overview of the basic Tali Forth 2 structure. For those
interested, a later chapter will provide far more detail.
