% Manual for Tali Forth for the 65c02
% Scot W. Stevenson <scot.stevenson@gmail.com>
% First version: 02. Mar 2018
% This version: 03. Mar 2018
% See:
% - https://en.wikibooks.org/wiki/LaTeX/Document_Structure#Document_classes
% - https://www.sharelatex.com/blog/2013/08/02/thesis-series-pt1.html
\documentclass[a4paper,notitlepage]{report}
\usepackage[width=150mm,top=25mm,bottom=25mm]{geometry}
\usepackage{fancyhdr}
\pagestyle{fancy}

\usepackage{graphicx}
\usepackage{makeidx}
\makeindex

\usepackage{lmodern}    % for typewriter
\usepackage{listings}
\lstset{basicstyle=\ttfamily,breaklines=true} % use monospace

\usepackage{color}
\usepackage[colorlinks=true]{hyperref}   % must be last

% Add real titlepage

\title{The Tali Forth 2 Manual}
\author{Scot W.~Stevenson}
\date{\today}


% --------------------------------
\begin{document}

\maketitle

\begin{abstract}
        Tali Forth 2 is a bare-metal ANSI(ish) Forth for the 65c02 8-bit MPU. 
        It aims to be, roughly in order of importance: 

        \begin{description}

        \item [Easy to try.]
                \href{https://github.com/scotws/TaliForth2}{Download the source}
                -- or even just the binary -- and you can immediately
                run it in an emulater. This lets you experiment with a
                working 8-bit Forth for the 65c02 without any special
                configuration.

        \item [Simple.] The subroutine-threaded (STC) design and happily
                overcommented source code give hobbyists the chance to study a
                working Forth at the lowest level. The manual -- this
                document -- explains structure and code in detail. The aim
                is to make it easy to port Tali Forth 2
                to various 65c02 hardware projects.

        \item [Specific.] Many Forths available are `general' implementations with
                a small core adapted to the target processor. Tali Forth 2 was
                written as a "bare metal Forth" for the 65c02 8-bit MPU
                and that MPU only, with its strengths and limitations in
                mind.

        \item [Standardized.] Most Forths available for the 65c02 are based on ancient,
                outdated templates such as FIG Forth. Learning Forth with them is like
                trying to learn modern English by reading Chaucer. Tali
                Forth (mostly) follows the current ANSI Standard.
\end{description}

        Tali Forth is hosted at GitHub at
        \href{https://github.com/scotws/TaliForth2}{https://github.com/scotws/TaliForth2}
        where the newest version can be found. The discussion thread is at
        6502.org at
        \href{http://forum.6502.org/viewtopic.php?f=9\&t=2926}{http://forum.6502.org/viewtopic.php?f=9\&t=2926}.
        
\end{abstract}

% --------------------------------
\tableofcontents
\listoffigures
\listoftables

% --------------------------------
\part{Introduction}

\chapter{The Why}
        \input{ch_why}

\chapter{Overview of Tali Forth}
        \input{ch_overview}

% --------------------------------
\part{User Guide}

\chapter{Installing}
        \input{ch_installation}

\chapter{Running}
        \input{ch_running}

\chapter{The Editor}
\textit{(Currently, there is no editor installed.)}


% --------------------------------
\part{Developer Guide}

\chapter{How Tali Forth works}
        \input{ch_internals}

\chapter{Developing}
        \input{ch_developing}

\chapter{Future plans}

% --------------------------------

\appendix

\chapter{FAQ}
        \input{ch_faq}

\chapter{Forth tests}
        \input{ch_tests}

\chapter{Thanks}
        \input{ch_thanks}

\input{bibliography}

% See https://en.wikibooks.org/wiki/LaTeX/Indexing#Sophisticated_indexing for
% details
\printindex

\end{document}
